Predicting Bike Usage at UofT St. George with Bayesian Poisson Regression

### 2024-11-29 paper/paper.qmd
dicussion updated
dicussion updated

The core of this study is to extract a sample of 27 bike-sharing stations within the University of Toronto campus from all bike-sharing usage data between January 1, 2017, and September 30, 2024. The total usage of bike-sharing at each station is calculated for each 4-hour interval. A Bayesian Poisson regression model is then employed to predict the usage based on the station, year, month, date, and the specific 4-hour interval of the day.

@@

## Estimand
This study aims to estimate the usage of a specific bike-sharing station within the University of Toronto campus during a particular 4-hour interval. By converting time into year, month, day, and the specific 4-hour interval of the day, it accounts for the overall trend, seasonal effects, and hourly variation in station usage. The core objective is to explore the temporal changes in the usage of 27 bike-sharing stations within the campus, thereby providing policymakers with recommendations for bike allocation to improve commuting efficiency on campus.

@@

@fig-daily_usage shows the daily usage totals of 27 shared bicycle stations from January 1, 2017, to September 30, 2024. It can be observed that the usage of the 27 bike-sharing stations at the University of Toronto's St. George campus has shown a steady upward trend, with an average increase of 1 use every 20 days. At the same time, there are clear seasonal variations: usage is higher in the spring and summer months (April to September) and significantly lower in the fall and winter months (October to March). Additionally, there was an abnormal surge in usage from June to September 2024, which can be attributed to two main reasons. First, historical data indicates that usage typically spikes in August and September each year, likely due to increased demand from students returning to campus for the start of the academic year. Second, during the same period in 2024, Bike Share Toronto added 52 new stations, most of which were located near the St. George campus, significantly enhancing the network's coverage and convenience[@bike]. These two factors combined led to the appearance of anomalies in August and September 2024.
```{r, fig.pos="H"}
#| label: fig-daily_usage
#| fig-cap: Daily usage from January 2017 to September 2024
#| echo: false
#| message: false
#| warning: false
# 计算线性模型
lm_model <- lm(total_count ~ time, data = daily_counts)

# 提取斜率
slope <- coef(lm_model)["time"]

@@

## What the Study Brought to the World
The study reveals a few key insights about bike-sharing usage on university campuses. Specifically, we learn that bike-sharing usage on the University of Toronto's St. George campus exhibits significant temporal patterns, influenced by season and time of day. For instance, usage drops significantly during winter months and peaks from 8 AM to 8 PM, with little demand between midnight and 8 AM. This indicates that bike-sharing systems need to adapt dynamically to changing demand patterns to be more effective, with an emphasis on considering seasonal and daily time variations for better resource allocation.

The findings provide valuable lessons for urban planners and campus management on improving bike-sharing infrastructure by optimizing the number and distribution of bikes according to the predicted demand in different seasons and time slots. This also underscores the broader principle that effective urban transport planning, even at a small scale like a university campus, requires a data-driven approach to match supply with temporal demand.

>>>>

The estimand of this study aims to predict the usage of a specific bike-sharing station on the University of Toronto campus during a particular 4-hour interval. By transforming time into components such as year, month, day, and the specific 4-hour interval of the day, the analysis incorporates overall trends, seasonal effects, and hourly variations in station usage. The primary objective is to investigate temporal changes in the usage of 27 bike-sharing stations on campus, thereby providing policymakers with recommendations for optimizing bike allocation to enhance commuting efficiency. To achieve this estimand, the study extracts a sample of 27 bike-sharing stations within the University of Toronto campus from all bike-sharing usage data recorded between January 1, 2017, and September 30, 2024. A Bayesian Poisson regression model is then employed to predict station usage based on factors such as station location, year, month, day, and the specific 4-hour interval of the day.

@@

@@

@fig-daily_usage shows the daily usage totals of 27 shared bicycle stations from January 1, 2017, to September 30, 2024. It can be observed that the usage of the 27 bike-sharing stations at the University of Toronto's St. George campus has shown a steady upward trend, with an average increase of 1 use every 20 days. At the same time, there are clear seasonal variations: usage is higher in the spring and summer months (April to September) and significantly lower in the fall and winter months (October to March). Additionally, there was an abnormal surge in usage from June to September 2024, which can be attributed to two main reasons. First, historical data indicates that usage typically spikes in August and September each year, likely due to increased demand from students returning to campus for the start of the academic year. Second, during the same period in 2024, Bike Share Toronto added 52 new stations, most of which were located near the St. George campus, significantly enhancing the network's coverage and convenience[@bike]. These two factors combined led to the appearance of anomalies in August and September 2024.
```{r, fig.pos="H"}
#| label: fig-daily_usage
#| fig-cap: Daily usage from January 2017 to September 2024
#| echo: false
#| message: false
#| warning: false

start_asdate <- start 
start_asdate$time <- as.Date(start_asdate$time)
daily_counts <- start_asdate %>%
  group_by(time) %>%
  summarise(total_count = sum(count))

lm_model <- lm(total_count ~ time, data = daily_counts)

slope <- coef(lm_model)["time"]

@@

## Optimization Suggestions for Campus Bike-Sharing Systems

This study identifies significant patterns in the use of bike-sharing systems on the University of Toronto’s St. George campus, including temporal, seasonal, and annual trends. The findings reveal that 79.8% of bike usage occurs between 8 AM and 8 PM, while demand between midnight and 8 AM is minimal. This highlights the strong alignment between bike usage and users’ daily routines. To improve commuting efficiency, it is recommended that the university collaborates with the bike-sharing provider to increase bike availability during peak hours (e.g., before and after class times) and optimize station layouts, particularly in high-demand areas such as dormitories, academic buildings, and transit hubs. During low-demand hours (e.g., early morning), maintenance and redistribution efforts can be scheduled to ensure the availability of bikes during peak periods.

Seasonal data also show that bike usage significantly decreases during winter, approaching zero, while increasing steadily during spring and summer, peaking in August and September during the start of the academic year. This suggests the need for proactive and dynamic adjustments in bike supply. During the start of the academic year, expanding station capacity and offering user-friendly navigation services can help new and returning students quickly familiarize themselves with the system. In winter, efforts should focus on improving station infrastructure (e.g., anti-freeze facilities) and maintaining safety under cold conditions. Additionally, by leveraging predictive analytics to forecast station-specific demand fluctuations, resources can be optimized in advance, minimizing inconveniences caused by bike shortages or overcrowded parking during peak times.

## This Study Provides a Research Methodology for Shared Bikes in Semi-Closed Areas

This study provides actionable recommendations for improving user experience in semi-closed environments like university campuses. Unlike urban open systems, bike usage in campuses is strongly influenced by fixed schedules such as class timings and campus events. This approach can be further extended to other similar semi-closed areas, such as industrial parks, manufacturing facilities, and residential complexes, where movement patterns are often dictated by predictable schedules, shifts, or local routines.

In these environments, bike-sharing systems can be optimized to align with peak demand periods, such as shift changes in factories, scheduled breaks in office parks, or recreational times in residential areas. By leveraging predictive analytics and data-driven insights, administrators can strategically allocate bikes and docking stations to match demand patterns, ensuring users have access when and where they need it most.

Moreover, this methodology is cost-effective. By targeting specific time windows and locations with higher demand, resources can be distributed more efficiently, reducing waste and operational costs. This reduces the need for large-scale deployment of bikes and infrastructure, which is often necessary in open urban systems. Instead, investments can be focused on key high-demand areas within these semi-closed environments, maximizing return on investment while enhancing user satisfaction.
